---
title: "STAT 415 Application 4"
author: "Andrew Kerr"
date: ''
format:
  html:
    number_sections: yes
    code-fold: true
    toc: true
    toc-title: "Outline"
    toc-depth: 3
embed-resources: true
---


# Instructions

This RMarkdown file provides a template for you to fill in.
Read the file from start to finish, completing the parts as indicated.
Some blank "code chunks" are provided; you are welcome to add more (CTRL-ALT-I) as needed.
There are also a few places where you should type text responses.
Feel free to add additional text responses as necessary.

# Submission instructions


When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.
When you are finished

- click **Knit** and check to make sure that you have no errors and everything runs properly. (Fix any problems and redo this step until it works.)
- Make sure your type your name(s) at the top of the notebook where it says "Type your name(s) here". If you worked in a team, you will submit a single notebook with both names; make sure both names are included
- Submit your completed file in Canvas.

# Set up

You might need a few R packages, which you can install using `install.packages`, e.g., `install.packages("janitor")`.


```{r}
library(knitr)

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```


```{r}
library(tidyverse)
library(viridis)
library(kableExtra)
library(janitor)

```

# Introduction

Does uniform color give athletes an advantage over their competitors?
To investigate this question, Hill and Barton (*Nature*, 2005) examined the records in the 2004 Olympic Games for four combat sports: boxing, tae kwon do, Greco-Roman wrestling, and freestyle wrestling.
*Competitors in these sports were randomly assigned to wear either a red or a blue uniform.*
The researchers investigated whether competitors wearing one color won significantly more often than those wearing the other color.
They collected data on a total of 457 matches.

Let the parameter $\theta$ represent the probability that a competitor wearing *red* wins a match.
Let $y$ be the number of matches, out of 457, in which the competitor wearing red defeats the competitor wearing blue.

(@) Assume a Beta(50, 50) prior for $\theta$.
Use simulation to approximate the prior predictive distribution of $y$ and plot it.
(It might make more sense to plot $y/457$.)
Explain why a Beta(50, 50) distribution seems like an unreasonable choice for the prior of $\theta$.

```{r}
# Enter your code here - add as many chunks as you want
```

**TYPE YOUR RESPONSES HERE.**

(@) Assume a Beta(200, 200) prior distribution for $\theta$.
Use simulation to approximate the prior predictive distribution of $y$ and plot it.
(It might make more sense to plot $y/457$.)
Explain why a Beta(200, 200) distribution seems like a more reasonable choice than Beta(50, 50) for the prior of $\theta$.


```{r}
# Enter your code here - add as many chunks as you want
```

**TYPE YOUR RESPONSES HERE.**


(@) Assume a Beta(50, 50) prior distribution for $\theta$.
The researchers analyzed a total of 457 matches, and they found that the competitor wearing red defeated the competitor wearing blue in 248 of these matches.
Identify the posterior distribution, its mean and SD, and compute *and interpret* central 50%, 80%, and 98% posterior credible intervals.
Create a plot showing prior, (scaled) likelihood, and posterior.


```{r}
# Enter your code here - add as many chunks as you want
```

**TYPE YOUR RESPONSES HERE.**


(@) Assume a Beta(200, 200) prior distribution for $\theta$.
The researchers analyzed a total of 457 matches, and they found that the competitor wearing red defeated the competitor wearing blue in 248 of these matches.
Identify the posterior distribution, its mean and SD, and compute *and interpret* central 50%, 80%, and 98% posterior credible intervals.
Create a plot showing prior, (scaled) likelihood, and posterior.


```{r}
# Enter your code here - add as many chunks as you want
```

**TYPE YOUR RESPONSES HERE.**


(@) How sensitive are the results to the choice of prior (Beta(50, 50) versus Beta(200, 200))?

**TYPE YOUR RESPONSES HERE.**


(@) Express the posterior mean as a weighted average of the prior mean and the sample proportion.
(Use the results based on the Beta(200, 200) prior.)


**TYPE YOUR RESPONSES HERE.**


(@) We might say that if the probability of winning by the competitor wearing red is between 0.49 and 0.51, then any color advantage is small enough to be considered unimportant.
In other words $\theta$ outside of 0.49 and 0.51 could be interpreted as one color giving a meaningful advantage.
Compute and interpret the posterior probability that $\theta$ is not in [0.49, 0.51].
(If you disagree with the range [0.49, 0.51] feel free to use different values.)



```{r}
# Enter your code here - add as many chunks as you want
```

**TYPE YOUR RESPONSES HERE.**


(@) Write 1-3 clearly worded sentences reporting the conclusions of your analysis in this context.


**TYPE YOUR RESPONSES HERE.**




(@) A Beta(200, 200) distribution is closely approximated by a Normal distribution with mean 0.5 and SD 0.025.
Suppose we had assumed a N(0.5, 0.025) prior for $\theta$, with the 248/457 sample data.
Use grid approximation to approximate the posterior distribution of $\theta$ and its mean and SD.
Create a plot showing prior, (scaled) likelihood, and posterior.
How does it compare to the posterior from the previous part (with the Beta(200, 200) prior)?

```{r}
# Enter your code here - add as many chunks as you want
```

**TYPE YOUR RESPONSES HERE.**


(@) Use simulation to approximation the posterior distribution of the odds $\phi = \frac{\theta}{1-\theta}$.
Find and interpret in context a central 98% credible interval for $\phi$.
(Use the posterior based on the Beta(200, 200) prior.)

```{r}
# Enter your code here - add as many chunks as you want
```

**TYPE YOUR RESPONSES HERE.**



