---
title: "STAT 415 Application 4"
author: "Andrew Kerr"
date: ''
format:
  html:
    number_sections: yes
    code-fold: true
    toc: true
    toc-title: "Outline"
    toc-depth: 3
embed-resources: true
---


# Instructions

This RMarkdown file provides a template for you to fill in.
Read the file from start to finish, completing the parts as indicated.
Some blank "code chunks" are provided; you are welcome to add more (CTRL-ALT-I) as needed.
There are also a few places where you should type text responses.
Feel free to add additional text responses as necessary.

# Submission instructions


When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.
When you are finished

- click **Knit** and check to make sure that you have no errors and everything runs properly. (Fix any problems and redo this step until it works.)
- Make sure your type your name(s) at the top of the notebook where it says "Type your name(s) here". If you worked in a team, you will submit a single notebook with both names; make sure both names are included
- Submit your completed file in Canvas.

# Set up

You might need a few R packages, which you can install using `install.packages`, e.g., `install.packages("janitor")`.


```{r}
library(knitr)

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```


```{r}
library(tidyverse)
library(viridis)
library(kableExtra)
library(janitor)

```

# Introduction

Does uniform color give athletes an advantage over their competitors?
To investigate this question, Hill and Barton (*Nature*, 2005) examined the records in the 2004 Olympic Games for four combat sports: boxing, tae kwon do, Greco-Roman wrestling, and freestyle wrestling.
*Competitors in these sports were randomly assigned to wear either a red or a blue uniform.*
The researchers investigated whether competitors wearing one color won significantly more often than those wearing the other color.
They collected data on a total of 457 matches.

Let the parameter $\theta$ represent the probability that a competitor wearing *red* wins a match.
Let $y$ be the number of matches, out of 457, in which the competitor wearing red defeats the competitor wearing blue.

(@) Assume a Beta(50, 50) prior for $\theta$.
Use simulation to approximate the prior predictive distribution of $y$ and plot it.
(It might make more sense to plot $y/457$.)
Explain why a Beta(50, 50) distribution seems like an unreasonable choice for the prior of $\theta$.

```{r}
set.seed(123)

sim_theta <- rbeta(10000, 50, 50)
sim_y <- rbinom(10000, 457, sim_theta)

ggplot() +
  geom_histogram(aes(x = sim_y/457, y = after_stat(density)), 
                 fill = "cornflowerblue",
                 bins = 45) +
  theme_bw() +
  labs(title = "Proportion of Red Wins (y/457) ~ Beta(50, 50)", x = "Proportion")
```

Beta(50, 50) seems unreasonable because of how much it varies. Assuming color has no effect, then we would predict theta to be 0.5. Although 0.5 is the center of this distribution, 0.4 and 0.6 are still very plausible.

(@) Assume a Beta(200, 200) prior distribution for $\theta$.
Use simulation to approximate the prior predictive distribution of $y$ and plot it.
(It might make more sense to plot $y/457$.)
Explain why a Beta(200, 200) distribution seems like a more reasonable choice than Beta(50, 50) for the prior of $\theta$.

```{r}
set.seed(123)

sim_theta <- rbeta(10000, 200, 200)
sim_y <- rbinom(10000, 457, sim_theta)

ggplot() +
  geom_histogram(aes(x = sim_y/457, y = after_stat(density)), 
                 fill = "cornflowerblue",
                 bins = 45) +
  theme_bw() +
  labs(title = "Proportion of Red Wins (y/457) ~ Beta(200, 200)", x = "Proportion")
```

Beta(200, 200) seems to be a more reasonable distribution because it has a smaller spread/variability. It is also centered at 0.5, with most of the plausibility bunched up at this center.

(@) Assume a Beta(50, 50) prior distribution for $\theta$.
The researchers analyzed a total of 457 matches, and they found that the competitor wearing red defeated the competitor wearing blue in 248 of these matches.
Identify the posterior distribution, its mean and SD, and compute *and interpret* central 50%, 80%, and 98% posterior credible intervals.
Create a plot showing prior, (scaled) likelihood, and posterior.

```{r}
# Beta prior
alpha_prior = 50
beta_prior = 50

# data
n = 457
y = 248

# Beta posterior
alpha_post = alpha_prior + y
beta_post = beta_prior + n - y

# scaled likelihood, depends on data: n, y
likelihood_scaled <- function(theta) {
  likelihood <- function(theta) {
    dbinom(x = y, size = n, prob = theta)
  }
  scaling_constant <- integrate(likelihood, lower = 0, upper = 1)[[1]]
  likelihood(theta) / scaling_constant
}

# Plot
ggplot(data.frame(x = c(0, 1)),
       aes(x = x)) +
  # prior
  stat_function(fun = dbeta,
                args = list(shape1 = alpha_prior,
                            shape2 = beta_prior),
                col = "cornflowerblue",
                linewidth = 1) +
  # (scaled) likelihood
  stat_function(fun = likelihood_scaled,
                col = "forestgreen",
                linewidth = 1) +
  # posterior
  stat_function(fun = dbeta,
                args = list(shape1 = alpha_post,
                            shape2 = beta_post),
                col = "firebrick",
                linewidth = 1) +
  labs(x = "theta",
       y = "") +
  theme_bw()
```

```{r}
post_mean <- alpha_post / (alpha_post + beta_post)
post_var <- (alpha_post * beta_post) / 
            ((alpha_post + beta_post)^2 * (alpha_post + beta_post + 1))
post_sd <- sqrt(post_var)

ci_50 <- qbeta(c(0.25, 0.75), alpha_post, beta_post)
ci_80 <- qbeta(c(0.10, 0.90), alpha_post, beta_post)
ci_98 <- qbeta(c(0.01, 0.99), alpha_post, beta_post)

cat("Posterior Mean:", post_mean, "\n")
cat("Posterior SD:", post_sd, "\n")
cat("50% Credible Interval:", ci_50, "\n")
cat("80% Credible Interval:", ci_80, "\n")
cat("98% Credible Interval:", ci_98, "\n")
```

- There is a 50% probability that the true proportion of red uniform wins lies between 0.52 and 0.55, given our data and prior beliefs.

- There is a 80% probability that the true proportion of red uniform wins lies between 0.51 and 0.56, given our data and prior beliefs.

- There is a 98% probability that the true proportion of red uniform wins lies between 0.49 and 0.58, given our data and prior beliefs.

(@) Assume a Beta(200, 200) prior distribution for $\theta$.
The researchers analyzed a total of 457 matches, and they found that the competitor wearing red defeated the competitor wearing blue in 248 of these matches.
Identify the posterior distribution, its mean and SD, and compute *and interpret* central 50%, 80%, and 98% posterior credible intervals.
Create a plot showing prior, (scaled) likelihood, and posterior.

```{r}
# Beta prior
alpha_prior = 200
beta_prior = 200

# data
n = 457
y = 248

# Beta posterior
alpha_post = alpha_prior + y
beta_post = beta_prior + n - y

# scaled likelihood, depends on data: n, y
likelihood_scaled <- function(theta) {
  likelihood <- function(theta) {
    dbinom(x = y, size = n, prob = theta)
  }
  scaling_constant <- integrate(likelihood, lower = 0, upper = 1)[[1]]
  likelihood(theta) / scaling_constant
}

# Plot
ggplot(data.frame(x = c(0, 1)),
       aes(x = x)) +
  # prior
  stat_function(fun = dbeta,
                args = list(shape1 = alpha_prior,
                            shape2 = beta_prior),
                col = "cornflowerblue",
                linewidth = 1) +
  # (scaled) likelihood
  stat_function(fun = likelihood_scaled,
                col = "forestgreen",
                linewidth = 1) +
  # posterior
  stat_function(fun = dbeta,
                args = list(shape1 = alpha_post,
                            shape2 = beta_post),
                col = "firebrick",
                linewidth = 1) +
  labs(x = "theta",
       y = "") +
  theme_bw()
```

```{r}
post_mean <- alpha_post / (alpha_post + beta_post)
post_var <- (alpha_post * beta_post) / 
            ((alpha_post + beta_post)^2 * (alpha_post + beta_post + 1))
post_sd <- sqrt(post_var)

ci_50 <- qbeta(c(0.25, 0.75), alpha_post, beta_post)
ci_80 <- qbeta(c(0.10, 0.90), alpha_post, beta_post)
ci_98 <- qbeta(c(0.01, 0.99), alpha_post, beta_post)

cat("Posterior Mean:", post_mean, "\n")
cat("Posterior SD:", post_sd, "\n")
cat("50% Credible Interval:", ci_50, "\n")
cat("80% Credible Interval:", ci_80, "\n")
cat("98% Credible Interval:", ci_98, "\n")
```

- There is a 50% probability that the true proportion of red uniform wins lies between 0.51 and 0.53, given our data and prior beliefs.

- There is a 80% probability that the true proportion of red uniform wins lies between 0.50 and 0.54, given our data and prior beliefs.

- There is a 98% probability that the true proportion of red uniform wins lies between 0.48 and 0.56, given our data and prior beliefs.

(@) How sensitive are the results to the choice of prior (Beta(50, 50) versus Beta(200, 200))?

The results are not very sensitive to the prior, both resulting in roughly the same posterior mean and SD.

(@) Express the posterior mean as a weighted average of the prior mean and the sample proportion.
(Use the results based on the Beta(200, 200) prior.)

```{r}
prior_mean <- alpha_prior / (alpha_prior + beta_prior)
tot <- alpha_post + beta_post

prior_mean * ((alpha_prior + beta_prior) / tot) + y / n * (n / tot)
```

(@) We might say that if the probability of winning by the competitor wearing red is between 0.49 and 0.51, then any color advantage is small enough to be considered unimportant.
In other words $\theta$ outside of 0.49 and 0.51 could be interpreted as one color giving a meaningful advantage.
Compute and interpret the posterior probability that $\theta$ is not in [0.49, 0.51].
(If you disagree with the range [0.49, 0.51] feel free to use different values.)

```{r}
prob_below <- pbeta(0.49, alpha_post, beta_post)
prob_above <- 1 - pbeta(0.51, alpha_post, beta_post)

prob_outside <- prob_below + prob_above

cat("Probability θ < 0.49:", prob_below, "\n")
cat("Probability θ > 0.51:", prob_above, "\n")
cat("Total P(θ outside [0.49, 0.51]):", prob_outside, "\n")
```

The probability that theta is not in the interval 0.49 to 0.51 is 80%. 

(@) Write 1-3 clearly worded sentences reporting the conclusions of your analysis in this context.

Using a prior of Beta(200, 200), we have a posterior mean of 0.52. If we assume that if the probability of a red uniform winning is between 0.49 and 0.51 is not a significant color effect, then with our data and prior we have an 80% probability of having a meaningful color effect. 

(@) A Beta(200, 200) distribution is closely approximated by a Normal distribution with mean 0.5 and SD 0.025.
Suppose we had assumed a N(0.5, 0.025) prior for $\theta$, with the 248/457 sample data.
Use grid approximation to approximate the posterior distribution of $\theta$ and its mean and SD.
Create a plot showing prior, (scaled) likelihood, and posterior.
How does it compare to the posterior from the previous part (with the Beta(200, 200) prior)?

```{r}
theta <- seq(0.4, 0.6, length.out = 1000)

# Data
n <- 457
y <- 248

# Normal prior: N(0.5, 0.025)
prior <- dnorm(theta, mean = 0.5, sd = 0.025)
scaled_prior <- prior / sum(prior)

# Likelihood: Binomial(n, y | theta)
likelihood <- dbinom(y, size = n, prob = theta)
scaled_likelihood <- likelihood / sum(likelihood)

# Posterior
posterior <- prior * likelihood
posterior <- posterior / sum(posterior)

ggplot(data.frame(theta = theta), aes(x = theta)) +
  # Prior (Normal)
  geom_line(aes(y = scaled_prior), color = "cornflowerblue", linewidth = 1) +
  # Scaled likelihood
  geom_line(aes(y = scaled_likelihood), color = "forestgreen", linewidth = 1) +
  # Posterior
  geom_line(aes(y = posterior), color = "firebrick", linewidth = 1) +
  labs(x = "theta", y = "Density", 
       title = "Normal Prior (Blue) vs. Likelihood (Green) vs. Posterior (Red)") +
  theme_bw()
```

```{r}
post_mean <- sum(theta * posterior)
post_sd <- sqrt(sum(theta^2 * posterior) - post_mean^2)

cat("Posterior Mean:", post_mean, "\n")
cat("Posterior SD:", post_sd, "\n")
```

The mean and SD are very close to being the same!

(@) Use simulation to approximation the posterior distribution of the odds $\phi = \frac{\theta}{1-\theta}$.
Find and interpret in context a central 98% credible interval for $\phi$.
(Use the posterior based on the Beta(200, 200) prior.)

```{r}
set.seed(123)

theta_samples <- rbeta(10000, alpha_post, beta_post)
phi_samples <- theta_samples / (1 - theta_samples)

ci_98 <- quantile(phi_samples, probs = c(0.01, 0.99))
cat("98% Credible Interval for ϕ:", ci_98, "\n")
```

The odds ratio of red uniforms have an advantage over blue is between 0.94 and 1.3 with 98% probability.
