---
title: "STAT 415 Application 9"
author: "Andrew Kerr"
date: ''
output:
  html_document:
    toc: yes
    toc_float: yes
    number_sections: yes
  pdf_document: default
embed-resources: true
---

# Introduction

In this application you will perform a Bayesian analysis comparing two population proportions.
You will also code an MCMC algorithm from scratch.


# Instructions

This RMarkdown file provides a template for you to fill in.
**Read the file from start to finish, completing the parts as indicated.**
**Some code is provided for you. Be sure to run this code, and make sure you understand what it's doing.**
In particular, mind the **PAUSE**s; be sure to **PAUSE** and think about these questions before proceeding.
You are welcome to copy and paste code from handouts or my online notes.
Be sure to include all code and relevant output in your submission.



You'll need a few R packages, which you can install using `install.packages`

```{r, warning = FALSE, message = FALSE}

library(tidyverse)
library(janitor)
library(viridis)
library(kableExtra)
library(ggmosaic)

library(brms)
library(bayesplot)
library(tidybayes)

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  eval = TRUE
)

```





# Submission instructions

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.
When you are finished

- Click **Knit** and check to make sure that you have no errors and everything runs properly.
(Fix any problems and redo this step until it works.)
- Make sure your type your name(s) at the top of the notebook where it says "Type your name(s) here".
If you worked in a team, you will submit a single notebook with both names; make sure all names are included.
- Submit your completed files in Canvas.




# Setup

Many studies of animals involve tagging the animals, but can the tags be harmful?
Specifically, are metal bands used for tagging penguins harmful?
[Researchers studied the effects of banding in a sample of 100 king penguins near Antarctica](https://www.nature.com/articles/nature09630) that had already been tagged with RFID (radio frequency identification) chips.
Of these 100 penguins, 50 were randomly assigned to receive metal bands on their flippers (in addition to the RFID chip), while the remaining 50 did not receive metal bands.
The penguins were followed for 4.5 years to see which penguins survived.


Let $\theta_1$  denote the probability that a penguin *without* a metal band *survives* a 4.5-year period, and let $\theta_2$ denote the probability that a penguin *with* a metal band *survives* a 4.5-year period.
(Throughout, "penguin" refers to whatever population of penguins this sample would be reasonably representative of.)

**PAUSE** to consider your prior distribution for $\theta_1$ and $\theta_2$.
Suppose you don't know much about penguin survival in Antarctica; what prior distribution might you assume?
Would you necessarily assume that $\theta_1$ and $\theta_2$ are independent?
Why might you want to assume some dependence between the parameters in the prior?
(We'll discuss some issues in choosing the prior later.)



# Sample data

Here's a summary of the sample data.

- Of the 50 penguins without metal bands, 31 survived this 4.5-year study period.
- Of the 50 penguins with metal bands, 16 survived the 4.5-year study period.


```{r}
data = read_csv(here::here("data", "penguins.csv"))

data |> head() |> kbl() |> kable_styling()
```


```{r}
table(data$Metal_band, data$Survival_status)
```



```{r}
ggplot(data) +
  geom_mosaic(aes(x = product(Survival_status, Metal_band),
                  fill = Survival_status)) +
  theme_mosaic()
```



We will use the sample data to perform Bayesian inference to compare the probability of surviving with and without the metal band.


# `brms` code


We'll start with a prior that assumes $\theta_1$ and $\theta_2$ are independent.
We'll choose the same marginal prior distribution for both $\theta_1$ and $\theta_2$ to reflect a prior belief of "no difference".
A quick Google search reveals that the lifespan of Antarctic king penguins is 15-20 years, though we don't know how old the penguins were at the start of the 4.5-year study.
Let's assume a prior with a mean that represents penguins are more likely to survive than not, but is still fairly non-informative.
Therefore, our prior assumes

- $\theta_1$ and $\theta_2$ are independent according to the prior
- $\theta_1$ has a Beta(3, 2) prior distribution
- $\theta_2$ has a Beta(3, 2) prior distribution



The following `brms` code approximates the posterior distribution given the sample data via simulation and summarizes it.

```{r}
fit <-
  brm(data = data, 
      family = bernoulli(link = identity),
      formula = Survival_status ~ 0 + Metal_band,
      prior(beta(3, 2), lb = 0, ub = 1),
      iter = 3500,
      warmup = 1000,
      chains = 4,
      refresh = 0)
```

```{r}
summary(fit)
```


```{r}
plot(fit)
```


```{r}
pairs(fit)
```



# Difference in proportions

1. Use the `brms` results to approximate the posterior distribution of $\theta_1 - \theta_2$.
Create a plot of the posterior distribution, and write a few sentences describing the posterior distribution in context.

```{r}
posterior = fit |>
  spread_draws(b_Metal_bandyes, b_Metal_bandno) |>
  rename(mu1 = b_Metal_bandno, mu2 = b_Metal_bandyes) |>
  mutate(mu_diff = mu1 - mu2)

ggplot(posterior,
       aes(x = mu_diff)) +
  geom_histogram(aes(y = after_stat(density)),
                 col = "cornflowerblue", fill = "white", bins = 100) +
  geom_density(linewidth = 1, col = "cornflowerblue") +
  labs(x = "Difference in means (probability)") +
  theme_bw()

posterior |>
  summarise(
    mean = mean(mu_diff),
    sd = sd(mu_diff)
  ) |> kbl() |> kable_styling()
```

The posterior distribution has a mean of 0.27 and SD of 0.09. In other words, on 
average the difference in survival probability for penguins without bands is 0.27
compared to those with bands. The distribution appears normal, with a slightly longer
left tail.

1. Compute and interpret in context (approximate) 50%, 80%, and 98% posterior credible intervals for $\theta_1 - \theta_2$.

```{r}
quantile(posterior$mu_diff,
         c(0.01, 0.10, 0.25, 0.5, 0.75, 0.90, 0.99))
```

There is a posterior probability of [50%/80%/98%] that the average probability of
survival for bandless penguins is between [0.21/0.16/0.07] and [0.34/0.39/0.47]
units greater than banded penguins.

1. Compute and interpret in context the (approximate) posterior probability that $\theta_1-\theta_2>0$.

```{r}
sum(posterior$mu_diff > 0) / length(posterior$mu_diff)
```

The posterior probability that bandless penguins live longer than banded penguins
is 0.9988.

# Relative risk

1. Use the `brms` results to approximate the posterior distribution of $\theta_1 / \theta_2$.
Create a plot of the posterior distribution, and write a few sentences describing the posterior distribution in context.
(Note: you might see a few super extreme ratios that will throw of the scale, so you might consider $\log(\theta_1 / \theta_2)$ instead for plotting.)

```{r}
posterior = fit |>
  spread_draws(b_Metal_bandyes, b_Metal_bandno) |>
  rename(mu1 = b_Metal_bandno, mu2 = b_Metal_bandyes) |>
  mutate(mu_ratio = mu1 / mu2)

ggplot(posterior,
       aes(x = log(mu_ratio))) +
  geom_histogram(aes(y = after_stat(density)),
                 col = "cornflowerblue", fill = "white", bins = 100) +
  geom_density(linewidth = 1, col = "cornflowerblue") +
  labs(x = "Log Difference in means (probability)") +
  theme_bw()

posterior |>
  summarise(
    mean = mean(mu_ratio),
    sd = sd(mu_ratio)
  ) |> kbl() |> kable_styling()
```

The mean and SD of the ratio are 1.86 and 0.41 respectively. This means that 
the probability of survival for bandless penguins is 1.86 times larger than that
for banded penguins (or 86% larger). The plot of the log ratio is slightly right 
skewed (and without log is has some extreme values to the right).

1. Compute and interpret in context (approximate) 50%, 80%, and 98% posterior credible intervals for $\theta_1 / \theta_2$.

```{r}
quantile(posterior$mu_ratio,
         c(0.01, 0.10, 0.25, 0.5, 0.75, 0.90, 0.99))
```

There is a posterior probability of [50%/80%/98%] that the average probability of
survival for bandless penguins is between [1.57/1.39/1.14] and [2.08/2.40/3.11]
times greater than that for banded penguins.

1. Compute and interpret in context the (approximate) posterior probability that $\theta_1/\theta_2>1$.

```{r}
sum(posterior$mu_ratio > 0) / length(posterior$mu_ratio)
```

The posterior probability that bandless penguins live longer than banded penguins
is 1.00.

# Conclusions

Is there evidence that penguins with metal bands are more likely to die than those without?
Can we conclude that the bands are the *cause* of any difference in survival probabilities?
Write a few sentences summarizing your analysis and reporting conclusions in context.
Be sure to support you conclusions with appropriate values reported in context.

There is evidence that penguins with metal bands are more likely to die than 
those without since both the difference and the ratio averages and credible
intervals were in favor of the bandless penguins living longer. We can conclude 
that the bands are the reason since we are assuming the populations are 
independent, and because we used random assignment when assigning bands to the
penguins. 

# MCMC 

Now you will write code to implement the Metropolis algorithm and run it to approximate the posterior distribution.
You don't have to get fancy about how the "steps" proceed; use whatever works for you: for loops, functions, replicate, vectorization, etc.
What's important is that you implement the proposal + acceptance/rejection properly.
In particular, make sure you know how data, likelihood, and prior enter in the picture.
Your code should take data, likelihood, and prior as input and return a simulated sample from the posterior distribution *without first computing the posterior.*

Given the current state, you can propose a new state using a "Bivariate Normal neighboorhood" like in the Steph Curry example.
But consider the scale of your parameters when choosing the $\delta$ parameter that determines the "neighborhood".


You are welcome to copy my code from the handouts as a starting point, but you will need to modify it to fit this situation.
You might also understand the algorithm better if you code it from scratch.

```{r}
pi_theta <- function(theta_1, theta_2) {
  if(theta_1 >= 0 || theta_1 <= 1 || theta_2 >= 0 || theta_2 <= 1){
      likelihood_theta_1 <- dbinom(31, size = 50, prob = theta_1)
      likelihood_theta_2 <- dbinom(16, size = 50, prob = theta_2)
      dbeta(theta_1, 3, 2) * dbeta(theta_2, 3, 2) * likelihood_theta_1 * likelihood_theta_2
    } else {
      0
    }
}

metropolis <- function(N, delta=c(0.05, 0.05), seed=123, output=F){
  set.seed(seed)
  results <- matrix(NA, nrow = N, ncol = 2)
  colnames(results) <- c("theta1", "theta2")
  
  cur_1 <- 0.5
  cur_2 <- 0.5
  results[1, ] <- c(cur_1, cur_2)
  
  for(i in 2:N){
    prop_1 <- cur_1 + rnorm(1, mean = 0, sd = delta[1])
    prop_2 <- cur_2 + rnorm(1, mean = 0, sd = delta[2])
    a <- min(1, pi_theta(prop_1, prop_2)/pi_theta(cur_1, cur_2))
    
    decision <- sample(c(F, T), size=1, prob=c(1-a, a))
    
    if(decision){
      cur_1 <- prop_1
      cur_2 <- prop_2
    }
    
    results[i, ] <- c(cur_1, cur_2)
    
    if(output){
      cat("Step:", i)
      cat("\nCurrent State:", cur_1, cur_2)
      cat("\nProposed State:", prop_1, prop_2)
      cat("\nAcceptance Proba.:", a)
      cat(
        "\nDecision:",
        if(decision){"Accepted!"}else{"Rejected!"}, 
        "\n------------------------------------\n"
        )
    }
    
  }
  return(results)
}
```

1. Code the algorithm, run it for 10 or so steps, and output the step-by-step results of the first 10 or so steps.
Do enough steps so that you see each of (1) a proposal that is automatically accepted, (2) a proposal that is accepted, but not automatically, (3) a proposal that is rejected.
For each step provide

    - the current state
    - the proposed state
    - the probability of accepting the proposed move
    - the decision to accept or reject the proposal

```{r}
results <- metropolis(10, seed=111, output=T)
```

1. Then simulate many steps of the chain, discard the first 1000 or so steps (warm up), and create

    - trace plots
    - a plot summarizing the joint posterior distribution of $(\theta_1, \theta_2)$ pairs
    - plots summarizing the marginal posterior distributions of $\theta_1$ and $\theta_2$ 
    - summaries of the posterior distributions (means, SDs, credible intervals, etc)

```{r}
results <- metropolis(10000, seed=111)

results <- results[-(1:1000), ]

par(mfrow = c(1, 2))
plot(1:nrow(results), results[,1], type = "l", xlab = "Step", ylab = "theta1")
plot(1:nrow(results), results[,2], type = "l", xlab = "Step", ylab = "theta2")

par(mfrow = c(1, 1))
ggplot(results, aes(theta1, theta2)) +
  geom_point(color = "cornflowerblue", alpha = 0.4) +
  stat_ellipse(level = 0.98, color = "black", size = 2) +
  stat_density_2d(color = "grey", size = 1) +
  theme_bw()

par(mfrow = c(1, 2))
hist(results[,1], breaks = 50, freq = FALSE,
 xlab = "theta", main = "Posterior Distribution theta1")
hist(results[,2], breaks = 50, freq = FALSE,
 xlab = "theta", main = "Posterior Distribution theta2")

as.data.frame(results) |>
  summarise(
    mean_theta1 = mean(theta1),
    mean_theta2 = mean(theta2),
    SD_theta1 = sd(theta1),
    SD_theta2 = sd(theta2)
  )

quantile(results[1,],
         c(0.01, 0.10, 0.25, 0.5, 0.75, 0.90, 0.99))
quantile(results[2,],
         c(0.01, 0.10, 0.25, 0.5, 0.75, 0.90, 0.99))
```

1. Compare the results from your MCMC algorithm to the `brms` output.
Are the results similar?

The results from the brms and my output are similar/agree that penguins without
bands have a greater probability of survival than those with bands. The mean and 
SD values are very similar, if not the same!


