---
title: "PP 05"
author: "Andrew Kerr"
format:
  html:
    number_sections: yes
    code-fold: true
    toc: true
    toc-title: "Outline"
    toc-depth: 3
embed-resources: true
---

```{r}
#| message: false
#| include: false 

library(tidyverse)
library(kableExtra)
```

Consider three tennis players Arya (“A”), Brienne (“B”), and Cersei (“C”). One of these players is better than the other two, who are equally good/bad. When the best player plays either of the others, she has a 2/3 probability of winning the match. When the other two players play each other, each has a 1/2 probability of winning the match. But you do not know which player is the best. Based on watching the players warm up, you start with subjective prior probabilities of 0.5 that A is the best, 0.35 that B is the best, and 0.15 that C is the best. A and B will play the first match, then A will play C in the second match, and B will play C in the third match.

Hint: the key to each of these problems is to correctly identify the likelihood.

Note: if you’ve taken a class with me before you might have seen a version of this problem. I encourage you to treat the problem as if you’ve never seen it before; don’t go looking for your old notes.

### 1. Suppose that A beats B in the first match. Construct a Bayes table to find your posterior probability that each of A, B, C is best given that A beats B in the first match. Compare the posterior probabilities to the prior probabilities. Explain how your probabilities changed, and why that makes sense.

```{r}
df <- data.frame(
  player = c("A", "B", "C"),
  prior = c(.5, .35, .15),
  likelihood_A_beats_B = c(2/3, 1/3, 1/2)
)

df <- df %>%
  mutate(
    product = prior * likelihood_A_beats_B,
    posterior = product / sum(product)
  )

df %>%
  kable(digits = 3)
```

The probability for player A increased since they won, while player Bs decreased since they lost. Player Cs very slightly decreased because they did not participate in the match, so we are still unsure about them.

### 2. Now suppose that after A beats B in the first match, then A beats C in the second match. Construct a Bayes table to find your posterior probability that each of A, B, C is best given the results of the first two matches. Use as the prior your posterior probabilities from the previous part. Explain how your probabilities changed, and why that makes sense.

```{r}
df <- df %>%
  mutate (
    prior = posterior,
    likelihood_A_beats_C = c(2/3, 1/2, 1/3) 
    ) %>%
  select(player, prior, likelihood_A_beats_C)

df <- df %>%
  mutate(
    product = prior * likelihood_A_beats_C,
    posterior = product / sum(product)
  )

df %>%
  kable(digits = 3)
```

Now the probability for player A is much greater than that of player B and C since player A has won against both of them. Meanwhile, the probabilities for players B and C have both decreased, more so that of player Cs since they just lost. The probability for player C is less than that for player B due to our initial prior which placed C lower than B.

### 3. Now suppose that after A beats B in the first match and A beats C in the second match, then B beats C in the third match. Construct a Bayes table to find your posterior probability that each of A, B, C is best given the results of the first three matches. Use as the prior your posterior probabilities from the previous part. Explain how your probabilities changed, and why that makes sense.

```{r}
df <- df %>%
  mutate (
    prior = posterior,
    likelihood_B_beats_C = c(1/2, 2/3, 1/3) 
    ) %>%
  select(player, prior, likelihood_B_beats_C)

df <- df %>%
  mutate(
    product = prior * likelihood_B_beats_C,
    posterior = product / sum(product)
  )

df %>%
  kable(digits = 3)
```

Player A did not change much because they did not player, while player B increased and C decreased since they won and lost respectively.

### 4. In the previous parts we updated posterior probabilities after each match. What if we waited until the results of all three matches? Construct a Bayes table to find your posterior probability that each of A, B, C is best given the results of the first three matches (A beats B, A beats C, B beats C). Use your original prior probabilities from the start of the problem. The likelihood should now reflect the results of the three matches. For example, if A is best, the likelihood that A beats B in the first match is 2/3, the likelihood that A beats C in the second match is 2/3, and the likelihood that B beats C in the third match is 1/2. If we assume the matches are independent^[1](https://canvas.calpoly.edu/courses/157089/assignments/1237284#fn1)^, then the likelihood of these results from the 3 matches if A is best is (2/3)(2/3)(1/2).

```{r}
df <- data.frame(
  player = c("A", "B", "C"),
  prior = c(.5, .35, .15),
  likelihood = c(
    (2/3)*(2/3)*(1/2), 
    (1/3)*(1/2)*(2/3), 
    (1/2)*(1/3)*(1/3)
    )
)

df <- df %>%
  mutate(
    product = prior * likelihood,
    posterior = product / sum(product)
  )

df %>%
  kable(digits = 3)
```

The final posterior is the same! It does not matter if you update it all at once or one at a time.
