---
title: "PP 08"
author: "Andrew Kerr"
format:
  html:
    number_sections: yes
    code-fold: true
    toc: true
    toc-title: "Outline"
    toc-depth: 3
embed-resources: true
---

```{r}
#| message: false
#| include: false 

library(tidyverse)
library(kableExtra)
```

Assume the amount of sleep (hours) that a randomly selected Cal Poly student gets on a typical week night follows a Normal distribution with mean theta hours and standard deviation 1.5 hours. 

Assume that the prior distribution of theta is Normal with mean 8 and standard deviation 0.5.

In a sample of 25 Cal Poly students, the sample mean amount of sleep is 7.1 hours.

You perform a Bayesian analysis of the data, and your posterior distribution is approximately Normal with posterior mean 7.34 and posterior standard deviation 0.26.

Note: you can use software to compute. However, you might want to compute by hand for practice (especially if you’re rusty on Normal distribution calculations).

### 1. There are a few SDs above. Explain in words in context what each of these SDs represent

1. 1.5: The variation in mean sleep student-to-student

2. 0.5: The variation of theta

3. 0.26: The variation of theta after taking into account the data

4. Also, what does 1.5 / sqrt(25) represent?: sample-to-sample variation of mean sleep over many samples of 25 students

### 2. Compute and interpret a 67% central prior credible interval for theta.

```{r}
qnorm(c(0.165, 1 - 0.165), 8, .5)
```

There is a prior credibility of 67% that the average hours of sleep of a CP student is between 7.5 and 8.5 hours.

### 3. Compute and interpret a 98% central prior credible interval for theta.

```{r}
qnorm(c(0.01, 1 - 0.01), 8, .5)
```

There is a prior credibility of 98% that the average hours of sleep of a CP student is between 6.8 and 9.1 hours.

### 4. Compute and interpret the prior probability that theta is less than 7.5.

```{r}
pnorm(7.5, 8, .5)
```

The prior probability that a CP students sleeps for less than 7.5 hours is about 16%.

### 5. Compute and interpret a 67% central posterior credible interval for theta.

```{r}
qnorm(c(0.165, 1 - 0.165), 7.34, .26)
```

There is a posterior credibility of 67% that the average hours of sleep of a CP student is between 7.0 and 7.6 hours.

### 6. Compute and interpret a 98% central posterior credible interval for theta.

```{r}
qnorm(c(0.01, 1 - 0.01), 7.34, .26)
```

There is a posterior credibility of 98% that the average hours of sleep of a CP student is between 6.7 and 7.9 hours.

### 7. Compute and interpret the posterior probability that theta is less than 7.5.

```{r}
pnorm(7.5, 7.34, .26)
```

The posterior probability that a CP students sleeps for less than 7.5 hours is about 73%.


