---
title: "PP 16"
author: "Andrew Kerr"
format:
  html:
    number_sections: yes
    code-fold: true
    toc: true
    toc-title: "Outline"
    toc-depth: 3
embed-resources: true
---

Heights (inches) of softball players follow a Normal distribution with mean $\theta$ and standard deviation 5.0. Assume a Normal(67.0, 2.0) prior distribution. In a random sample of 10 softball players the sample mean height is 64.0.

The endpoints of each of the 95% posterior credible interval and the 95% posterior prediction interval are of the form mean±2×SD

Choices for mean: 64.0, 64.1, 65.1, 66.8, 67.0, 67.2

Choices for SD: 0.2, 1.2, 2.0,, 3.0, 5.0, 5.2

1. For the posterior credible interval: identify from the lists above the most appropriate values of mean and SD and explain your choices without doing any calculations.

mean: 65.1 since it is between the prior and sample mean
sd: 1.2 since we will always decrease the SD with more data

2. For the posterior prediction interval: identify from the lists above the most appropriate values of mean and SD and explain your choices without doing any calculations.

mean: 65.1 for the same reason as above
sd: 5.2 since we know that the SD of heights is 5.0, and the prediction interval is larger

3. Identify the posterior distribution of $\theta$, including computing the values of any relevant parameters. Verify that the computations match your choices above.

```{r}
# Given values
mu_0 <- 67.0    # Prior mean
tau_0 <- 2.0    # Prior SD
prec_0 <- 1 / tau_0^2 # Prior precision
x_bar <- 64.0   # Sample mean
n <- 10         # Sample size
sigma <- 5.0    # Known SD of data
prec <- n / sigma^2 # Sample precision

# Posterior precision
prec_n <- prec_0 + prec

# Posterior mean
mu_n <- mu_0 * (prec_0 / (prec_0 + prec)) + x_bar * (prec / (prec_0 + prec))

# Posterior SD
tau_n <- sqrt(1 / prec_n)

cat("Posterior mean (μ_n):", mu_n, "\n")
cat("Posterior SD (τ_n):", tau_n, "\n")
```

4. Write a clearly worded sentence interpreting the posterior credible interval in context.

```{r}
c(mu_n - 1.96 * tau_n, mu_n + 1.96 * tau_n)
```

95% of the population of softball player heights is between 62.7 and 67.6 inches.

5. Use simulation to approximate the posterior predictive distribution; approximate its mean, SD, and describe its shape. You already know the posterior distribution, so you can simulate $\theta$ directly from its posterior distribution.

```{r}
set.seed(123)
sim <- rnorm(10000, mu_n, tau_n)

hist(sim)

cat("Posterior predictive mean:", mean(sim), "\n")
cat("Posterior predictive SD:", sd(sim), "\n")
```

Its normal!

6. It can be shown that in this situation that the variance of the posterior predictive distribution is the sum of \(\sigma^2\) and the posterior variance. Compute that value and compare to the simulated value from the previous part.

```{r}
sigma^2 + tau_n
var(sim)
```

It doenst?

7. Write a clearly worded sentence interpreting the posterior prediction interval in context.

```{r}
c(mean(sim) - 1.96 * sd(sim), mean(sim) + 1.96 * sd(sim))
```

95% of a randomly selected softball players height will fall between 62.7 and 67.6 inches.
